A Splitting Solution of the Linearity Challenge

### 2025-12-11 paper/Context.lagda.tex
snapshot

\begin{code}[hide]
{-# OPTIONS --rewriting #-}
open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
open import Relation.Binary.PropositionalEquality using (_≡_; refl; cong; cong₂)
open import Data.List.Base using (List; []; _∷_; [_]; _++_; map)

@@

We are going to adopt a nameless representation of channels whereby a channel is
identified by its position in a typing context. This representation is akin to
using De Bruijn indices, except that the index, instead of being represented
explicitly by a natural number, is computable from the \emph{proof} that the
type of the channel belong to the typing context.
%
Typing contexts are represented using \emph{lists} of types, where the
(polymorphic) \AgdaDatatype{List} type is defined in Agda's standard library.

@@

The most important operation concerning typing contexts is \emph{splitting}. The
splitting of $\ContextC$ into $\ContextD$ and $\ContextE$, which is denoted by
$\Splitting\ContextC\ContextD\ContextE$, represents the fact that $\ContextC$
contains all the types contained in $\ContextD$ and $\ContextE$, preserving both
their overall \emph{multiplicity} and also their relative \emph{order} within
$\ContextD$ and $\ContextE$. A \emph{proof} of
$\Splitting\ContextC\ContextD\ContextE$ shows how the types in $\ContextC$ are
distributed in $\ContextD$ and $\ContextE$ from left to right.

\begin{code}[hide]
infix 4 _≃_+_ _∋_⊳_
\end{code}
\begin{code}
data _≃_+_ : Context → Context → Context → Set where
  •   : [] ≃ [] + []
  <_  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ A ∷ Δ + Θ
  >_  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ Δ + A ∷ Θ
\end{code}

When splitting a context $\Context$ into $\ContextD + \ContextE$, for each type
in $\Context$ we use one of the prefix operators \AgdaInductiveConstructor{<}
and \AgdaInductiveConstructor{>} to indicate whether the type is meant to be
placed in $\ContextD$ or $\ContextE$, respectively.
%
As an example, below is the splitting of the context $[A, B, C, D]$ into $[B] +
[A, C, D]$.

@@

Context splitting enjoys a number of expected properties. In particular, it is
easy to see that splitting is commutative and that the empty context is both a
left and right unit of splitting.

@@

It is often the case that the context $\ContextD$ in a splitting
$\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. We use this
particular splitting to describe the presence of a channel of type $A$ in
$\ContextC$ and giving a name $\ContextE$ to the residual context. For this
particular case, we introduce the dedicated notation
$\SimpleSplitting\ContextC{A}\ContextE$ which we find clearer.

\begin{code}
_∋_⊳_ : Context → Type → Context → Set
Γ ∋ A ⊳ Δ = Γ ≃ [ A ] + Δ
\end{code}
\begin{code}[hide]
+-empty-l : ∀{Γ Δ} → Γ ≃ [] + Δ → Γ ≡ Δ
+-empty-l • = refl
+-empty-l (> p) = cong (_ ∷_) (+-empty-l p)
\end{code}

Note that, when specifying a splitting $\SimpleSplitting\ContextC{A}\ContextE$,
it may be convenient to use \AgdaFunction{≫} as a shortcut for a sequence of
applications of \AgdaInductiveConstructor{>} after the $A$ type has been
reached. For instance, \AgdaFunction{splitting-example₁} can be written
equivalently and in a more compact way as shown below. More usages of
\AgdaFunction{≫} will be given in \Cref{sec:examples}.

\begin{code}[hide]
module _ where
\end{code}
\begin{code}
  splitting-example₂ : (A ∷ B ∷ C ∷ D ∷ []) ∋ B ⊳ (A ∷ C ∷ D ∷ [])
  splitting-example₂ = > < ≫
\end{code}

To conclude the modeling of typing contexts, we define a predicate $\AgdaDatatype{Un}~\Context$ that holds
whenever $\Context$ is \emph{unrestricted}, namely when it is made of types of
the form $?A$. We will need this predicate in the definition of a server, which
must obey the typing rule \ServerRule of \Cref{tab:typing-rules}.

\begin{code}
data Un : Context → Set where
  un-[]  : Un []
  un-∷   : ∀{A Γ} → Un Γ → Un (`? A ∷ Γ)
\end{code}

% It is a simple fact that the merging of two unrestricted contexts
% $\Context_1$ and $\Context_2$ results in an unrestricted context.\Luca{Serve?}

\begin{code}[hide]
+-un : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → Un Γ₁ → Un Γ₂ → Un Γ
+-un • un-[] un-[] = un-[]
+-un (< p) (un-∷ un₁) un₂ = un-∷ (+-un p un₁ un₂)
+-un (> p) un₁ (un-∷ un₂) = un-∷ (+-un p un₁ un₂)
\end{code}

>>>>

\begin{code}[hide]
{-# OPTIONS --rewriting #-}
open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
open import Data.List.Base using (List; []; _∷_; [_]; _++_)
open import Relation.Unary
open import Relation.Binary.PropositionalEquality using (_≡_; refl; cong)

@@

We are going to adopt a nameless representation of channels whereby a channel is
identified by the position of its type in a typing context. This representation
is akin to using De Bruijn indices except that the index, instead of being
represented explicitly by a natural number, is computable from the proof that
the type of the channel belongs to the typing context.
%
Typing contexts are represented using lists of types, where the (polymorphic)
\AgdaDatatype{List} type and its constructors \AgdaInductiveConstructor{[]} and
\AgdaInductiveConstructor{\_∷\_} are defined in the module
\AgdaModule{Data.List} of Agda's standard library. We will keep using
$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts, even
though in the Agda formalization they are lists instead of finite maps.

@@

The most important operation concerning typing contexts is \emph{splitting}. The
splitting of $\ContextC$ into $\ContextD$ and $\ContextE$, which we denote by
$\Splitting\ContextC\ContextD\ContextE$, represents the fact that $\ContextC$
contains all the types contained in $\ContextD$ and $\ContextE$, preserving both
their overall \emph{multiplicity} and also their relative \emph{order} within
$\ContextD$ and $\ContextE$. A \emph{proof} of
$\Splitting\ContextC\ContextD\ContextE$ shows how the types in $\ContextC$ are
distributed in $\ContextD$ and $\ContextE$ from left to right.

\begin{code}[hide]
infix  4 _≃_+_
infixr 8 _─∗_
infixr 9 _∗_
\end{code}
\begin{code}
data _≃_+_ : Context → Context → Context → Set where
  •   : [] ≃ [] + []
  <_  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ A ∷ Δ + Θ
  >_  : ∀{A Γ Δ Θ} → Γ ≃ Δ + Θ → A ∷ Γ ≃ Δ + A ∷ Θ
\end{code}

When splitting a context $\Context$ into $\ContextD + \ContextE$, for each type
in $\Context$ we use one of the prefix operators \AgdaInductiveConstructor{<}
and \AgdaInductiveConstructor{>} to indicate whether the type is meant to be
placed in $\ContextD$ or $\ContextE$.
%
For example, below is a proof of the splitting $\Splitting{[A, B, C,
D]}{[B]}{[A, C, D]}$.

@@

Context splitting enjoys a number of expected properties. In particular, it is
easy to see that splitting is commutative and that the empty context/list is
both a left and right unit of splitting.

@@

% It is often the case that the context $\ContextD$ in a splitting
% $\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. We use this
% particular splitting to describe the presence of a channel of type $A$ in
% $\ContextC$ and giving a name $\ContextE$ to the residual context. For this
% particular case, we introduce the dedicated notation
% $\SimpleSplitting\ContextC{A}\ContextE$ which we find clearer.

% \begin{code}
% _∋_⊳_ : Context → Type → Context → Set
% Γ ∋ A ⊳ Δ = Γ ≃ [ A ] + Δ
% \end{code}
\begin{code}[hide]
+-empty-l : ∀{Γ Δ} → Γ ≃ [] + Δ → Γ ≡ Δ
+-empty-l • = refl
+-empty-l (> p) = cong (_ ∷_) (+-empty-l p)
\end{code}

Note that, when specifying a splitting $\Splitting\ContextC{[A]}\ContextE$, it
may be convenient to use \AgdaFunction{≫} as a shortcut for a sequence of
applications of \AgdaInductiveConstructor{>} after the $A$ type has been
reached. For instance, \AgdaFunction{splitting-example₁} can be written
equivalently and in a more compact way as shown below. More usages of
\AgdaFunction{≫} will be given in \Cref{sec:examples}.

\begin{code}[hide]
module _ where
\end{code}
\begin{code}
  splitting-example₂ : (A ∷ B ∷ C ∷ D ∷ []) ≃ [ B ] + (A ∷ C ∷ D ∷ [])
  splitting-example₂ = > < ≫
\end{code}

\begin{table}
  \begin{tabular}{|lll|}
    \hline
    \textbf{Notation} & \textbf{Definition} & \textbf{Meaning} \\
    \hline
    $\AgdaFunction{∀[}~P~\AgdaFunction]$ & $\forall\{x\} \to P~x$ & implicit universality \\
    $P~\AgdaFunction{⇒}~Q$ & $\lambda x \to P~x \to Q~x$ & implication \\
    $P~\AgdaFunction{∪}~Q$ & $\lambda x \to P~x~\AgdaDatatype{⊎}~Q~x$ & disjunction \\
    $P~\AgdaFunction{∩}~Q$ & $\lambda x \to P~x~\AgdaDatatype{×}~Q~x$ & conjunction \\
    $f~\AgdaFunction{⊢}~P$ & $\lambda x \to P~(f~x)$ & update \\
    $\AgdaFunction{U}$ & $\lambda x \to \AgdaModule{Data.Unit}.\AgdaDatatype{⊤}$ & universal
    set \\
    $\AgdaFunction{⋂[}~X~\AgdaFunction{:}~A~\AgdaFunction]~P$ & $\lambda x \to (X : A) \to P~X~x$ & infinitary conjunction \\
    \hline
  \end{tabular}
  \caption{\label{tab:predicates} Useful definitions from Agda's \AgdaModule{Relation.Unary} module.}
\end{table}

In the following we will make extensive use of predicates over contexts. For
this reason, it is worth recalling a number of definitions from the
\AgdaModule{Relation.Unary} module in Agda's standard library in
\Cref{tab:predicates}.

Context splitting is going to be pervasive in the subsequent development, to the
point that it contributes substantially to the complexity of the formalization.
For this reason, we invest some effort in defining a few abstractions that allow
us to occasionally hide context splittings and enhance code readability.
%
Inspired by separation logic~\cite{Hoare} and
following~\citet{RouvoetPoulsenKrebbersVisser20}, we define the \emph{separating
conjunction} $P \sand Q$ of two predicates $P$ and $Q$ over contexts:
\begin{code}
data _∗_ (P Q : Pred Context _) (Γ : Context) : Set where
  _⟨_⟩_ : ∀{Δ Θ} → P Δ → Γ ≃ Δ + Θ → Q Θ → (P ∗ Q) Γ
\end{code}

The predicate $P \sand Q$ holds for those contexts $\ContextC$ that can be split
into $\ContextD$ and $\ContextE$ where $P$ holds for $\ContextD$ and $Q$ holds
for $\ContextE$.
%
The constructor \AgdaInductiveConstructor{\_⟨\_⟩\_} has three explicit arguments
witnessing the splitting $\Splitting\ContextC\ContextD\ContextE$ along with
$P~\ContextD$ and $Q~\ContextE$.\Luca{Nei related fare presente che queste
definizioni seguono sostanzialmente \citet{RouvoetPoulsenKrebbersVisser20} ed
evidenziare le differenze.}

Along with \AgdaDatatype{∗} we define the \emph{separating implication} (also
known as ``magic wand'')
\begin{code}
_─∗_ : Pred Context _ → Pred Context _ → Context → Set
(P ─∗ Q) Δ = ∀{Θ Γ} → Γ ≃ Δ + Θ → P Θ → Q Γ
\end{code}
and prove that \AgdaFunction{─∗} can be used to curry \AgdaDatatype{∗}:
\begin{code}
curry∗ : ∀{P Q R} → ∀[ P ∗ Q ⇒ R ] → ∀[ P ⇒ Q ─∗ R ]
curry∗ F px σ qx = F (px ⟨ σ ⟩ qx)
\end{code}

To conclude the modeling of typing contexts, we define a predicate
\AgdaDatatype{Un} that holds for \emph{unrestricted} contexts, namely those
contexts solely made of types of the form $?A$. We will need this predicate in
the definition of a server, which must obey the typing rule \ServerRule of
\Cref{tab:typing-rules}.

\begin{code}
data Un : Context → Set where
  un-[]  : Un []
  un-∷   : ∀{A} → ∀[ Un ⇒ (`? A ∷_) ⊢ Un ]
\end{code}

It is easy to prove that $\Splitting\ContextC\ContextD\ContextE$ is unrestricted
if so are $\ContextD$ and $\ContextE$:

\begin{code}
∗-un : ∀[ Un ∗ Un ⇒ Un ]
\end{code}
\begin{code}[hide]
∗-un (un-[] ⟨ • ⟩ un-[])    = un-[]
∗-un (un-∷ un ⟨ < σ ⟩ un′)  = un-∷ (∗-un (un ⟨ σ ⟩ un′))
∗-un (un ⟨ > σ ⟩ un-∷ un′)  = un-∷ (∗-un (un ⟨ σ ⟩ un′))
\end{code}

### 2025-12-11 paper/Permutations.lagda.tex
snapshot

The definition of \AgdaDatatype{\_↭\_} data type is nearly the same found in
Agda's standard library. We have preferred defining our own notion of
permutation for improved simplicity and convenience: the
\AgdaInductiveConstructor{swap} constructor does not need a sub-permutation for
the tail of the typing context, which can always be performed, if needed, using
\AgdaInductiveConstructor{prep} and \AgdaInductiveConstructor{trans}. Also, and
more importantly, the data type is \emph{monomorphic} (it does not need to
relate lists of arbitrary type) and the quantified arguments $A$ and $B$ of
\AgdaInductiveConstructor{swap} and \AgdaInductiveConstructor{prep} are
implicit, which streamlines their usage in the rest of the code.

@@

↭solo-inv : ∀{A Γ} → [ A ] ↭ Γ → Γ ≡ [ A ]
↭solo-inv refl = refl
↭solo-inv (prep π) rewrite ↭empty-inv π = refl
↭solo-inv (trans π π′) rewrite ↭solo-inv π | ↭solo-inv π′ = refl

↭rot : ∀{A B C Γ} → (A ∷ B ∷ C ∷ Γ) ↭ (C ∷ A ∷ B ∷ Γ)
↭rot = trans (prep swap) swap
\end{code}

Just like permutations from Agda's standard library, also \AgdaDatatype{\_↭\_}
enjoys a number of expected properties, such as symmetry. Among them, in the
following we will make use of a specific property that relates context
splitting, list concatenation and permutations: if
$\Splitting\ContextC\ContextD\ContextE$, then $\ContextC$ is a permutation of
the concatenation of $\ContextD$ and $\ContextE$.

@@

↭solo : ∀{A Γ Γ′ Δ} → Γ ↭ Δ → Γ ∋ A ⊳ Γ′ → ∃[ Δ′ ] (Δ ∋ A ⊳ Δ′ × Γ′ ↭ Δ′)
↭solo π p with ↭split π p
... | _ , _ , q , π₁ , π₂ rewrite ↭solo-inv π₁ = _ , q , π₂

>>>>

The definition of \AgdaDatatype{\_↭\_} data type is nearly the same found in the
module \AgdaModule{Data.List.Relation.Binary.Permutation.Propositional} of
Agda's standard library. We have preferred defining our own notion of
permutation for improved simplicity and convenience: the
\AgdaInductiveConstructor{swap} constructor does not need a sub-permutation for
the tail of the typing context, which can always be performed, if needed, using
\AgdaInductiveConstructor{prep} and \AgdaInductiveConstructor{trans}. Also, and
more importantly, the data type is \emph{monomorphic} (it does not need to
relate lists of arbitrary type) and the quantified arguments $A$ and $B$ of
\AgdaInductiveConstructor{swap} and \AgdaInductiveConstructor{prep} are
implicit, which streamlines their usage in the rest of the code.

@@

↭solo-inv : ∀{A Γ} → [ A ] ↭ Γ → Γ ≡ [ A ]
↭solo-inv refl = refl
↭solo-inv (prep π) rewrite ↭empty-inv π = refl
↭solo-inv (trans π π′) rewrite ↭solo-inv π | ↭solo-inv π′ = refl
\end{code}

Just like permutations in Agda's standard library, also \AgdaDatatype{\_↭\_}
enjoys a number of expected properties, such as symmetry. Among them, in the
following we will make use of a specific property that relates context
splitting, list concatenation and permutations: if
$\Splitting\ContextC\ContextD\ContextE$, then $\ContextC$ is a permutation of
the concatenation of $\ContextD$ and $\ContextE$.

@@

↭solo : ∀{A Γ Γ′ Δ} → Γ ↭ Δ → Γ ≃ [ A ] + Γ′ → ∃[ Δ′ ] (Δ ≃ [ A ] + Δ′ × Γ′ ↭ Δ′)
↭solo π p with ↭split π p
... | _ , _ , q , π₁ , π₂ rewrite ↭solo-inv π₁ = _ , q , π₂

### 2025-12-11 paper/Process.lagda.tex
snapshot

\begin{code}[hide]
{-# OPTIONS --rewriting #-}
open import Data.Unit using (tt)
open import Data.Sum
open import Data.Product using (_×_; _,_; ∃; ∃-syntax)
open import Data.List.Base using (List; []; _∷_; [_]; _++_)
open import Relation.Unary
open import Relation.Binary.PropositionalEquality using (_≡_; refl; cong; cong₂)

@@

We adopt a representation of processes that is \emph{intrinsically typed} and
based on \emph{co-de Bruijn indexes}. The intrinsically typed representation
makes sure that only well-typed processes can be constructed. This choice
increases the effort in the definition of the datatypes for representing
processes and their operational semantics, but pays off in the rest of the
formalization for at least three reasons:
\begin{itemize}
\item we can adopt a \emph{nameless} representation of channels, thus avoiding
  all the technicalities and pitfalls that a named representation entails;
\item we conflate processes and typing rules in the same datatype, thus reducing
  the overall number of datatypes defined in the formalization;
\item the typing preservation results are embedded in the very definition of the
  operational semantics of processes and need not require separate proofs
  (\Cref{sec:agda-semantics}).
\end{itemize}

We use \emph{co-de Bruijn indexes}~\cite{McBride18} to represent channels. Co-de
Bruijn indexes are akin to de Bruijn indexes, with the key difference that the
typing context used to type a certain process $P$ is guaranteed to only contain
the type of those channels that actually occur in $P$. For this reason, typing
contexts are split as soon as possible, according to the structure of processes,
to make sure that channels are appropriately (and above all linearly)
distributed among sub-processes, so that each channel occurs (and is used)
exactly once.\footnote{Strictly speaking, the typing context of a
\AgdaInductiveConstructor{fail} process may contain channels that do not occur
(and therefore are not used) by the process, but we consider this case as an
exception that does not alter in a substantial way the provided intuition
underlying the use of co-de Bruijn representation.}

From this informal introduction to the representation of processes it is clear
that context splitting is going to be a pervasive notion in the subsequent
development, to the point that it contributes substantially to the complexity of
the formalization. For this reason, it is worth investing some effort in a
defining an abstraction that allows us to express the type of processes in a
more compact and hopefully more readable way.
\citet{RouvoetPoulsenKrebbersVisser20} observe that both channels and processes
can be thought of as \emph{predicates on typing contexts}:
\begin{itemize}
\item a channel of type $A$ is a predicate that holds for the singleton context
  $\AgdaFunction[~A~\AgdaFunction]$;
\item a process that is well typed in a typing context $\Context$ is a predicate
  that holds for $\Context$.
\end{itemize}

Inspired by separation logic~\cite{Hoare},
\citet{RouvoetPoulsenKrebbersVisser20} define the \emph{separating conjunction}
$P \sand Q$ of two predicates $P$ and $Q$. The separating conjunction of $P$ and
$Q$ holds for those contexts $\ContextC$ that can be split into $\ContextD$ and
$\ContextE$ where $P$ holds for $\ContextD$ and $Q$ holds for $\ContextE$. In
Agda, \citet{RouvoetPoulsenKrebbersVisser20} define $\sand$ as a record, thus:
\begin{code}
record _∗_ (P Q : Pred Context _) (Γ : Context) : Set where
  constructor _⟨_⟩_
  field
    {Δ Θ} : Context
    px    : P Δ
    p     : Γ ≃ Δ + Θ
    qx    : Q Θ
\end{code}
\begin{code}[hide]
infixr 9 _∗_
\end{code}

The record has five fields: the (implicit) fields $\ContextD$ and $\ContextE$,
which represent the decomposition of $\ContextC$; the fields \AgdaField{px} and
\AgdaField{qx}, which are proofs witnessing that $P$ and $Q$ hold respectively
for $\ContextD$ and $\ContextE$; the field \AgdaField{p}, which witnesses that
$\ContextD$ and $\ContextE$ are indeed a splitting of $\ContextC$. The
constructor \AgdaInductiveConstructor{$\_\langle\_\rangle\_$} provides a
convenient syntax for specifying the (explicit) fields of the record to provide
(and pattern match) separating conjunctions.

To illustrate a typical usage pattern of separating conjunction we first
introduce our first predicate \AgdaDatatype{Ch} on typing contexts, which we use
to describe well-typed channels:

@@

The predicate $\AgdaDatatype{Ch}~A$ holds for the singleton typing context
$[A]$. Hence the predicate $\AgdaDatatype{Ch}~A\sand\AgdaDatatype{Ch}~B$ holds
for those typing contexts that can be split into $[A]$ and $[B]$, as exemplified
below:
\begin{code}
_ : ∀{A B} → (Ch A ∗ Ch B) (A ∷ B ∷ [])
_ = ch ⟨ < > • ⟩ ch

_ : ∀{A B} → (Ch A ∗ Ch B) (B ∷ A ∷ [])
_ = ch ⟨ > < • ⟩ ch
\end{code}

\begin{table}
  \begin{tabular}{|lll|}
    \hline
    \textbf{Notation} & \textbf{Definition} & \textbf{Meaning} \\
    \hline
    $\AgdaFunction{∀[}~P~\AgdaFunction]$ & $\forall\{x\} \to P~x$ & implicit universality \\
    $P~\AgdaFunction{⇒}~Q$ & $\lambda x \to P~x \to Q~x$ & implication \\
    $P~\AgdaFunction{∪}~Q$ & $\lambda x \to P~x~\AgdaDatatype{⊎}~Q~x$ & disjunction \\
    $P~\AgdaFunction{∩}~Q$ & $\lambda x \to P~x~\AgdaDatatype{×}~Q~x$ & conjunction (non-separating) \\
    $f~\AgdaFunction{⊢}~P$ & $\lambda x \to P~(f~x)$ & update \\
    $\AgdaFunction{U}$ & $\lambda x \to \AgdaModule{Data.Unit}.\AgdaDatatype{⊤}$ & universal
    set \\
    $\AgdaFunction{⋂[}~X~\AgdaFunction{:}~A~\AgdaFunction]~P$ & $\lambda x \to (X : A) \to P~X~x$ & infinitary conjunction \\
    \hline
  \end{tabular}
  \caption{\label{tab:predicates} Used definitions from Agda's \AgdaModule{Relation.Unary} module.}
\end{table}

We now illustrate the datatype \AgdaDatatype{Proc} for representing processes
and then describe its key aspects in detail. This datatype uses, in addition to
$\sand$, a number of definitions from the \AgdaModule{Relation.Unary} module in
Agda's standard library which are worth recalling in \Cref{tab:predicates}.

@@

Let us start from the constructors representing input/output actions, namely
those from \AgdaInductiveConstructor{fail} through
\AgdaInductiveConstructor{contract}. They all basically follow the same pattern:
they possibly quantify over some types $A$ and $B$ and then (implicitly) over a
typing context $\Context$ through the function \AgdaFunction{∀[\_]} applied to a
predicate of the form $\cdots~\AgdaFunction{⇒}~\AgdaDatatype{Proc}$. The
predicate states how to build a process of a particular form, provided that the
typing context $\Context$ satisfies the predicate on the l.h.s. of
\AgdaFunction{⇒}.

@@

The constructor \AgdaInductiveConstructor{close} requires the typing context to
be the singleton list $[\One]$.

>>>>

\begin{code}[hide]
{-# OPTIONS --rewriting #-}
open import Data.Unit using (tt)
open import Data.Sum
open import Data.Product using (_,_)
open import Data.List.Base using ([]; _∷_; [_])
open import Relation.Unary

@@

We adopt an \emph{intrinsically-typed} representation of processes with
\emph{nameless} channels. The intrinsically-typed representation makes sure that
only well-typed processes can be constructed. This choice increases the effort
in the definition of the datatypes for representing processes and their
operational semantics, but pays off in the rest of the formalization for at
least three reasons:
\begin{itemize}
\item we need not give explicit names to channels, thus we avoid all the
  technicalities and pitfalls that a named representation entails;
\item we conflate processes and typing rules in the same datatype, thus reducing
  the overall number of datatypes defined in the formalization;
\item the typing preservation results are embedded in the very definition of the
  operational semantics of processes and need not require separate proofs
  (\Cref{sec:agda-semantics}).
\end{itemize}

Channels are not given any name. Instead, they are represented as terms
witnessing that their type is present in the typing context. This is known as
\emph{co-de Bruijn syntax}~\cite{McBride18}, according to which the typing
context associated with a term (a process, a channel) only contains the types of
the channels that actually occur within the term. For this reason, typing
contexts are split as soon as possible, according to the structure of processes,
to make sure that channels are appropriately (and above all linearly)
distributed among sub-processes, so that each channel occurs (and is used)
exactly once.\footnote{Strictly speaking, the typing context of a
\AgdaInductiveConstructor{fail} process may contain channels that do not occur
(and therefore are not used) by the process, but we consider this case as an
exception that does not alter in a substantial way the provided intuition
underlying the use of co-de Bruijn representation.}
%
Concretely, a channel of type $A$ is a predicate that holds for the singleton
context $\AgdaFunction[~A~\AgdaFunction]$:

@@

A process that is well typed in a typing context $\Context$ is a predicate that
holds for $\Context$. Here is the datatype \AgdaDatatype{Proc} for representing
processes:

@@

All the constructors representing input/output actions, namely those from
\AgdaInductiveConstructor{fail} through \AgdaInductiveConstructor{contract},
basically follow the same pattern: they possibly quantify over some types $A$
and $B$ and then (implicitly) over a typing context $\Context$ through the
function \AgdaFunction{∀[\_]} applied to a predicate of the form
$P~\AgdaFunction{⇒}~\AgdaDatatype{Proc}$. The predicate states how to build a
process that is well typed in $\Context$, provided that $\Context$ satisfies
$P$. In general, $P$ is a (separating) conjunction of sub-predicates
corresponding to the channel on which the process is acting and to the premises
of its typing rule.

@@

The constructor \AgdaInductiveConstructor{wait} corresponds to a process of the
form $\Wait\x.P$ and requires $\Context$ to (separately) satisfy
$\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊥}$, that is the channel $x$ on
which the process is operating, as well as \AgdaDatatype{Proc}, that is the
continuation process $P$, which must be well typed in the remaining typing
context.

The constructor \AgdaInductiveConstructor{close}, which corresponds to a process
of the form $\Close\x$, requires the typing context to be the singleton list
$[\One]$.

### 2025-12-11 paper/Type.lagda.tex
snapshot

At this stage we are finally ready to define the type of closed pre-types, those
having no free type variables. We call \AgdaDatatype{Type} closed pre-types.
Hereafter, we will seldom use pre-types again.

>>>>

At last we are ready to define the type of closed pre-types, those
having no free type variables. We call \AgdaDatatype{Type} closed pre-types.
From now on, we will seldom use pre-types again.
